---
title: "Getting started with genoPlotR"
output:
 rmarkdown::html_vignette:
  toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with genoPlotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

&NewLine;  
&NewLine;  

# Introduction

The **genoPlotR** package is intended to produce publication-grade graphics 
of gene and genome maps. With the amazing speed of data production of new 
DNA sequencing techniques and the increase in the number of software 
available to compare these sequences, there is a great need to graphically 
represent these sequences and their comparisons. A number of packages 
already exist (Artemis, ACT, mauve), but none of them produce easily 
reproducible, publication-grade graphics. The goal of this package is 
to fill in that gap. 

This vignette provides an introduction to **genoPlotR**, providing the user 
with examples of increasing complexity. It is not meant as a comprehensive
guide to all the functions and options of the package, but rather as 
a first approach to the package.

To load the library in an R session, type:

```{r setup, message=FALSE}
library(genoPlotR)
```

&NewLine;  
&NewLine;  

# Quick start

Loading the simplest dataset, applying a color scheme and some limits to the 
plotting area, adding a tree and some annotations. For more details about
that plot, refer to the first of the examples shown below.

```{r quick_start_data}
data(three_genes)
dna_segs <- three_genes$dna_segs
comparisons <- three_genes$comparisons
comparisons[[1]]$col <- apply_color_scheme(c(0.6, 0.4, 0.5), "grey")
names <- c("Huey", "Dewey", "Louie")
names(dna_segs) <- names
tree <- ade4::newick2phylog("(((Huey:4.2,Dewey:3.9):3.1,Louie:7.3):1);")
mid_pos <- middle(dna_segs[[1]])
xlims <- list(c(Inf, -Inf), c(-Inf, Inf), c(1850, 2800))
annot <- annotation(x1 = c(mid_pos[1], dna_segs[[1]]$end[2]),
                    x2 = c(NA, dna_segs[[1]]$end[3]),
                    text = c(dna_segs[[1]]$name[1], "region1"),
                    rot = c(30, 0),
                    col = c("blue", "black"))
```

Now plotting these three segments:

```{r quick_start_plot, fig.width=7, fig.height=2.5}
plot_gene_map(dna_segs = dna_segs, comparisons = comparisons,
              annotations = annot, annotation_height = 1.3,
              tree = tree, tree_width = 2,
              xlims = xlims,
              main = "Comparison of Huey, Dewey and Louie")
```

&NewLine;  
&NewLine;  

# Getting help

A lot of examples and help are available in the main functions, i.e. the 
reading functions (the various `read_dna_seg_from*` and 
`read_comparison_from*` functions) and the main plotting 
function, `plot_gene_map()`.

```{r getting_help3, eval=FALSE}
help("read_dna_seg")
help("read_comparison")
help("plot_gene_map")
```

Additionally, the web pages on 
[R-forge](http://genoplotr.r-forge.r-project.org/)
and [GitHub](https://github.com/MVPuijk/genoPlotR)
provide ways to get in touch with the **genoPlotR** community and to submit 
bugs and feature requests.

&NewLine;  
&NewLine;  

# Objects in **genoPlotR**

This section will give an overview of the different types of **R** objects in
**genoPlotR**.

## dna_seg

A `dna_seg` object is a collection of genes or elements along a 
genome, to be represented on a map. 

`dna_seg` objects need to have 4 columns, `name`, `start`, 
`end` and `strand`. Extra columns with the names `col`, `fill`,
`lty`, `lwd`, `pch`, `cex`, `gene_type`, and `region_type` will 
be used in the plotting process. Other extra columns will be kept in 
the object, but are not used.

```{r dna_seg}
names1 <- c("feat1", "feat2", "feat3")
starts1 <- c(2, 1000, 1050)
ends1 <- c(600, 800, 1345)
strands1 <- c("-", -1, 1)
cols1 <- c("blue", "grey", "red")
df1 <- data.frame(name = names1, start = starts1, end = ends1,
                  strand = strands1, col = cols1)

dna_seg1 <- dna_seg(df1)
str(dna_seg1)
```

## comparison

A `comparison` is a collection of similarities, representing the
comparison between two DNA segments. 

Objects (`data.frame`, `data.table`, or `list`) should have at least named
elements `start1`, `end1`, `start2` and `end2`. 
In addition, an optional `col` column can be added, that will give
the color of each comparison. Additional numeric columns can be used 
for automatic color-coding (via `gradient_color_scheme()`).

```{r comparison}
starts1 <- c(2, 1000, 1050)
ends1 <- c(600, 800, 1345)
starts2 <- c(50, 800, 1200)
ends2 <- c(900, 1100, 1322)
comparison1 <- as.comparison(data.frame(start1 = starts1, end1 = ends1,
                                        start2 = starts2, end2 = ends2))
str(comparison1)
```

## annotation

An `annotation` object is used to annotate a DNA segment. It has 
labels attached to positions. Each label can be attached to a single 
position or to a range. 

```{r annotation}
mid_pos <- middle(dna_segs[[1]])
annot1 <- annotation(x1 = mid_pos, text = dna_segs[[1]]$name)
str(annot1)
``` 

## tree

A tree description in Newick format can be parsed using the `ade4` package.
```{r tree}
tree <- ade4::newick2phylog("(((A_aaa:4.2,B_bbb:3.9):3.1,C_ccc:7.3):1);")
str(tree$leaves)
``` 

&NewLine;  
&NewLine;  

# Reading data

## DNA segments

Several formats can be read by **genoPlotR** to produce `dna_seg` objects:

* GenBank files (`read_dna_seg_from_genbank()`)
* EMBL files (`read_dna_seg_from_embl()`)
* FASTA files (`read_dna_seg_from_fasta()`)
* PTT (protein table) files, tabular versions of GenBank files 
(`read_dna_seg_from_ptt()`)
* User generated tabular files (`read_dna_seg_from_tab()`)

The function `read_dna_seg_from_file()` is a wrapper function, that
will attempt to guess the correct format of the file.

The first three files are common biological formats and can be downloaded 
from major databases, such as the [NCBI](https://www.ncbi.nlm.nih.gov/)
and the [ENA](https://www.ebi.ac.uk/ena/browser/home). The 
definition of EMBL and GenBank files can be found
[here](https://ftp.ebi.ac.uk/pub/databases/embl/doc/FT_current.html).

## Comparisons

**genoPlotR** can read tabular files, either user-generated tab files 
(`read_comparison_from_tab()`), or from BLAST output
(`read_comparison_from_blast()`). To produce files that are readable by 
**genoPlotR**, BLAST+ commands should be run using the `outfmt -6` or
`outfmt -7` options.

## Mauve output

The backbone output of the 
[Mauve genome aligner](https://darlinglab.org/mauve/mauve.html) can be parsed 
using `read_mauve_backbone()` (Tested with Mauve 2.3.1).

The function will return a list consisting of a list of `dna_seg` objects 
and the corresponding `comparisons`.

&NewLine;  
&NewLine;  

# Plotting data

There is only one plotting function in **genoPlotR**, `plot_gene_map()`.
Many arguments are available, but here is a list of the most important ones.
Check the documentation for a more thorough description.

:::: {style="display: grid; display: grid; grid-template-columns: 1fr 9fr; grid-column-gap: 10px;"}

::: {}
**`dna_segs`**
:::
::: {}
A list of DNA segment objects.  
:::

::: {}
**`comparisons`**
:::
::: {}
A list of comparisons. Should contain one element less than `dna_segs`.  
:::

::: {}
**`tree`**
:::
::: {}
A phylogenetic tree to be plotted at the left of the figure.  
:::

::: {}
**`annotations`**
:::
::: {}
An annotation object, or a list of annotations. Will display annotations to 
the first, or to all DNA segments, respectively.
:::

::: {}
**`xlims`**
:::
::: {}
A list of even-numbered numeric vectors, giving the borders of sub-segments to 
be plotted. The vector `c(1,5000,8000,6000)` will display 
two sub-segments (1 to 5000 and 6000 to 8000), the second being in reverse 
orientation. 
:::

::: {}
**`main`**
:::
::: {}
A title to the plot.
:::

::: {}
**`scale`**
:::
::: {}
Should a scale be displayed at the bottom right of the plot?
:::

::: {}
**`dna_seg_scale`**
:::
::: {}
Allows to control the addition of scales to each segments. If simply
`TRUE`, will display a scale on each segment. If a vector, a scale will 
be displayed for the corresponding `TRUE` element.
:::

::: {}
**`global_color_scheme`**
:::
::: {}
Generate a new set of colors for the DNA segments and/ or comparisons.
:::

::: {}
**`plot_new`**
:::
::: {}
Turn off to avoid creating a new plot. Especially useful to integrate
a **genoPlotR** plot in a larger figure.
:::

::::

&NewLine;  
&NewLine;  

# Other useful functions

`gradient_color_scheme()`  Makes a set of colors, either a gray scale or shades
of red and blue, that can then be applied to `comparisons`.  
`uniform_color_scheme()`  Applies a color scheme to `dna_segs` and/ or 
`comparisons`.  
`middle()`  Useful to get the middle of a gene, especially to
create annotations.  
`auto_annotate()`  Automatically generates annotations from `dna_segs`.  
`edit_dna_segs()`  Uses a data frame to edit a list of `dna_segs`.
`data()` Type `data(package = "genoPlotR")` to get the full list of datasets.  

&NewLine;  
&NewLine;  

# Examples

This section gives step-by-step examples, of gradual complexity. The last
one shows how to combine several plots and to annotate already generated 
plots.

For all examples, the first step is to load the library.

```{r examples, message=FALSE}
library(genoPlotR)
```

## Example 1: A very simple example

The data used in this example is a dataset included in **genoPlotR**, but
for the sake of demonstration, it will be recreated from scratch. To retrieve
this dataset from the package, use `data(three_genes)`.

### Data generation and a very simple plot

First, three `dna_segs` are generated, from data frames with columns 
`name`, `start`, `end`, `strand` and `col`.

```{r ex1_make_dna_segs}
df1 <- data.frame(name = c("feat1", "feat2", "feat3"),
                  start = c(2, 1000, 1050),
                  end = c(600, 800, 1345),
                  strand = c(-1, -1, 1),
                  col = c("blue", "grey", "red"))
dna_seg1 <- dna_seg(df1)
df2 <- data.frame(name = c("feat1", "feat2", "feat3"),
                  start = c(50, 800, 1200),
                  end = c(900, 1100, 1322),
                  strand = c(-1, 1, 1),
                  col = c("blue", "grey", "red"))
dna_seg2 <- dna_seg(df2)
df3 <- data.frame(name = c("feat1", "feat2", "feat3"),
                  start = c(1899, 2108, 2803),
                  end = c(2034, 2732, 3620),
                  strand = c(-1, -1, 1),
                  col = rep("blue", 3))
dna_seg3 <- dna_seg(df3)
dna_segs <- list(dna_seg1, dna_seg2, dna_seg3)
```

Then, create two `comparison` objects from data frames with columns
`start1`, `end1`, `start2`, `end2` and `col`. In the first comparison, the
starts and ends correspond to the genes created above.

```{r ex1_make_comparisons}
df4 <- data.frame(start1 = dna_seg1$start,
                  end1 = dna_seg1$end,
                  start2 = dna_seg2$start,
                  end2 = dna_seg2$end)
comparison1 <- comparison(df4)
df5 <- data.frame(start1 = c(50, 800),
                  end1 = c(500, 1100),
                  start2 = c(1899, 2732),
                  end2 = c(2034, 2508),
                  col = c("#67000D", "#08306B"))
comparison2 <- comparison(df5)
comparisons <- list(comparison1, comparison2)
```

This constitutes all the material required to create a basic plot.

```{r ex1_basic_plot, fig.width=6, fig.height=2.5}
plot_gene_map(dna_segs = dna_segs, comparisons = comparisons)
```
  
### A more elaborate plot

Many options can be added to this simple plot. To start, the first 
comparison will be colored in grey scale, using an arbitrary scale passed 
to the function `gradient_color_scheme()`.

```{r ex1_gradient}
comparisons[[1]]$col <- gradient_color_scheme(c(0.6, 0.4, 0.5), "grey")
```

Second, names and a phylogenetic tree are prepared. The function 
`newick2phylog()` from the package `ade4` is used to prepare an
object suitable for **genoPlotR**. The elements of the list of 
`dna_segs` must be named and correspond to the labels of the
tree object, to avoid confusion.

```{r ex1_tree}
names <- c("Huey", "Dewey", "Louie")
names(dna_segs) <- names
tree_HDL <- ade4::newick2phylog("(((Huey:4.2,Dewey:3.9):3.1,Louie:7.3):1);")
```

Third, annotations to the first segment are added. The first gene is annotated
in its middle (thus the use of the `middle()` function), while a region 
comprising the second and third gene are annotated with a square bracket.
Note the use of a numeric value for `x1` and `NA` for the first
gene, and the use of two numeric values for the second region.

```{r ex1_annots}
mid_pos <- middle(dna_segs[[1]])
annot <- annotation(x1 = c(mid_pos[1], dna_segs[[1]]$end[2]),
                    x2 = c(NA, dna_segs[[1]]$end[3]),
                    text = c(dna_segs[[1]]$name[1], "region1"),
                    rot = c(30, 0),
                    col = c("grey", "black"))
```

Finally, the plot is drawn, adding a title, allowing 2 inches for the tree
width, and changing the annotation space to 1.3 lines. The different options 
can be checked by removing/adding the different arguments to `plot_gene_map()`. 

```{r ex1_adv_plot, fig.width=6, fig.height=2.5}
plot_gene_map(dna_segs = dna_segs, comparisons = comparisons,
              annotations = annot, annotation_height = 1.3,
              tree = tree_HDL, tree_width = 2,
              main = "Comparison of Huey, Dewey and Louie")
```

To explore the different options, new windows can be opened by using 
`x11()` or `window()`, depending on the OS, or saved to files 
using for example `png()` or `pdf()`. To finish the plot, the 
function `dev.off()` should be called.

The plot can be saved to a file by first adding a call to one of the graphical
device functions of **R** (e.g. `pdf()`, `png()`, `jpeg()`), then making the 
call(s) to `plot_gene_map()`, and finally the function `dev.off()`.

## Example 2: Generating data online

This section will give an example of how data can be retrieved and generated
online, using the resources available at the 
[NCBI](http://www.ncbi.nlm.nih.gov/). This example was devised on 5/3/2010,
and given the relatively rapid rate of changes in public databases, it is 
possible that part or all of this example becomes obsolete. Please contact
the author of this document (lionel.guy@ebc.uu.se) in such a case.

In this example, the genomes of *Bartonella henselae* 
(RefSeq accession NC_005956) and *Bartonella quintana* (RefSeq 
accession NC_005955) will be compared.

### Retrieving dna_seg data

To retrieve GenBank files, the NCBI Nucleotide Database can be directly 
queried, provided that the accession number is known. In such a case, the
database can be accessed from the home page of
[NCBI](http://www.ncbi.nlm.nih.gov/), by entering the accession number 
(e.g. NC_005956) in the search field and by selecting "Nucleotide" as the
database. The GenBank file is then displayed. It can be saved in 
a suitable format by clicking on "Send to:", selecting "Complete Record" and
"File", and selecting "GenBank (full)" as the format.

If the accession number is unknown, the 
[NCBI genome database](https://www.ncbi.nlm.nih.gov/datasets/genome/)
can be queried for the name of the organism. 
From the list of assemblies, a GenBank format file can be downloaded by 
selecting "Download" from the "Action" column, and selecting the 
"Sequence and annotation (GBFF)" file type. 

Do this for both NC_005956 and NC_005955.

### Performing web-based BLAST and retrieving result

The two previously retrieved genomes can be compared with BLAST, which can be
done locally or online. The online version of BLAST can be found on the NCBI
website [here](https://blast.ncbi.nlm.nih.gov/Blast.cgi). From there, the 
genomes can be aligned by entering `NC_005956` as the query sequence, selecting
"Align two or more sequences", and then entering `NC_005955` as the subject
sequence. The other arguments to BLAST can then be modified at will to suit 
the needs of the comparison. Once the BLAST search is run, the hit table 
(which can be read by **genoPlotR**) can be downloaded by clicking on
"Download" and then "Hit Table(text)" in the result page.

If BLAST+ is installed locally and findable by R, then you can do this from 
within R. The function `comparisons_from_dna_segs()` makes a list of
`comparisons` by running BLAST first, and making FASTA files from GenBank files
as necessary. Alternatively, DIAMOND can be chosen as well using the argument
`tool = "diamond"`.

```{r ex_2_load_offline, results='hide', eval=FALSE}
BH_vs_BQ <-  comparisons_from_dna_segs(files = c("NC_005956.gb", "NC_005955.gb"),
                                       tool = "blast", algorithm = "blastn")
```

### Plotting

Provided that the GenBank files and their comparison have been saved
under the correct names in the directory where R was started, 
they can be parsed directly by **genoPlotR**.

```{r ex_2_load_online, results='hide', eval=FALSE} 
BH <- read_dna_seg_from_file("NC_005956.gb")
BQ <- read_dna_seg_from_file("NC_005955.gb")
BH_vs_BQ <- read_comparison_from_blast("NC_005956_NC_005955")
```

Alternatively, `data(barto)` contains `dna_segs` made from these two genomes,
as well as a `comparison` that was made from BLAST results.

```{r ex2_load_offline}
data(barto)
BH <- barto$dna_segs[[3]]
BQ <- barto$dna_segs[[4]]
BH_vs_BQ <- barto$comparisons[[3]]
```

The data can now be plotted, using the `xlims` argument to plot only the first
50,000 base pairs.

```{r ex2_plot, fig.width=6, fig.height=1.8}
xlims <- list(c(1,50000), c(1,50000))
plot_gene_map(dna_segs = list(BH, BQ),
              comparisons = list(BH_vs_BQ),
              dna_seg_labels = c("BH", "BQ"),
              xlims = xlims,
              main = "BH vs BQ, comparison of the first 50 kb",
              gene_type = "side_blocks",
              alpha_comparisons = 0.5,
              global_color_scheme = "gradient",
              dna_seg_scale = TRUE, scale = FALSE)
```

## Example 3: Mauve alignment of four *Bartonella* genomes

**genoPlotR** is able to parse the backbone file produced by Mauve (for details
on the format, see <https://darlinglab.org/mauve/user-guide/files.html>).

The elements of the DNA segments are no longer genes, but Mauve blocks. 
Similarly, the comparison reflects the correspondences between these blocks. 
The strand indicates the orientation of the block with respect to a reference,
which by default is the first genome in the comparison. A single list is 
returned, which contains two lists, one containing the `dna_segs` and
the other one the `comparisons`.

The genomes in the backbone file are not named, so it is advised to be 
cautious in the order of the names given.

In this example, 4 genomes of *Bartonella* have been compared with 
Mauve 2.3.1. The smaller blocks (smaller than 10 kb) are filtered out, and
the second genome (which is the largest) is taken as the reference. 
The dataset can also be accessed in the package by running
`data(mauve_bbone)`. 

```{r ex3_read_data}
bbone_file <- system.file('extdata/barto.backbone', package = 'genoPlotR')
bbone <- read_mauve_backbone(bbone_file, ref = 2, filter_low = 10000)
names <- c("B_bacilliformis", "B_grahamii", "B_henselae", "B_quintana")
names(bbone$dna_segs) <- names
```

The lengths of the comparisons can be calculated by adding the length on both 
sides of the comparisons.

By adding `global_color_scheme = "gradient"` and 
`color_scheme_column = "length"`
Now plotting, using `global_color_scheme()` to color the segments 
according to their lengths.


```{r ex3_length}
for (i in 1:length(bbone$comparisons)) {
  cmp <- bbone$comparisons[[i]]
  bbone$comparisons[[i]]$length <- 
    abs(cmp$end1 - cmp$start1) + abs(cmp$end2 - cmp$start2)
}
```

We can then color the comparisons based on these calculated lengths using the
`global_color_scheme` and `color_scheme_column` arguments.

```{r ex3_plot, fig.width=7, fig.height=2}
plot_gene_map(dna_segs = bbone$dna_segs, 
              comparisons = bbone$comparisons,
              global_color_scheme = c("length", "increasing", "red_blue", 0.7),
              override_color_schemes = TRUE)
```

## Example 4: Several sub-segments of four *Bartonella* genomes

This examples presents how to use the `xlims` argument to represent 
several sub-segments of the same segment, and how to show them in a reverse
orientation.

The data used here is also a comparison between four *Bartonella* genomes
(see above), but the comparison has been performed using BLAST. The dataset
is available in the package (`data(barto)`).

First, the data is loaded and a tree is created. 

```{r ex4_load_tree}
data(barto)
tree_barto <- ade4::newick2phylog("(BB:2.5,(BG:1.8,(BH:1,BQ:0.8):1.9):3);")
```

The `xlims` argument is then created. It is a list with as many elements
as there are `dna_segs`. Each element is an even-numbered numeric 
vector, containing the left and right border of each sub-segment, 
consecutively. In the first DNA segment, two sub-segments are shown: between
1,415,000 and 1,445,000 in a reverse orientation, and between 1,380,000 and
1,412,000, in "normal" orientation. In the remaining DNA segments, 3, 3, and 1 
subsegments are shown, respectively, all in normal orientation.

```{r ex4_xlims}
xlims <- list(c(1445000, 1415000, 1380000, 1412000),
              c(  10000,   45000,   50000,   83000, 90000, 120000),
              c(  15000,   36000,   90000,  120000, 74000,  98000),
              c(   5000,   82000))
```

To complete the example, annotations are added. Genes that have a gene name 
(i.e. these for which the name is not the locus or synonym name) are used
to annotate the segment. Only every fourth gene is annotated, to avoid
overlapping the tags. The function `middle()` is used to retrieve the 
middle of each element of the DNA segments.

```{r ex4_annots}
annots <- lapply(barto$dna_segs, function(x) {
  mid <- middle(x)
  annot <- annotation(x1 = mid, text = x$name, rot = 30)
  idx <- grep("^[^B]", annot$text, perl = TRUE)
  annot[idx[idx %% 4 == 0], ] 
})
```

Finally, plot the result, using scales on each DNA segment, adding a title,
and not limiting the plotting area to the longest segment, to allow for a 
better placement of the sub-segments.

```{r ex4_plot, fig.width=6, fig.height=4}
plot_gene_map(barto$dna_segs, barto$comparisons, tree = tree_barto,
              annotations = annots,
              xlims = xlims,
              limit_to_longest_dna_seg = FALSE,
              dna_seg_scale = TRUE, scale = FALSE,
              main = "Comparison of homologous segments in 4 Bartonella genomes")
```

## Example 5: Two segments of the Y chromsome in human and chimp

In this example, the ability to plot introns and exons is demonstrated.

First, data is loaded. The dataset included in the package is used, but any
GenBank file containing introns and exons can be used. Some annotations are
added: for each segment, the range of each gene is calculated, and 
corresponding annotations are created. 

```{r ex5_load_annot}
data(chrY_subseg)
genes_homo <- unique(chrY_subseg$dna_segs[[1]]$gene)
x_homo <- sapply(genes_homo, function(x)
                 range(chrY_subseg$dna_segs[[1]]
                       [chrY_subseg$dna_segs[[1]]$gene == x,])
                 )
annot_homo <- annotation(x1 = x_homo[1,], x2 = x_homo[2,],
                         text = dimnames(x_homo)[[2]])
genes_pan <- unique(chrY_subseg$dna_segs[[2]]$gene)
x_pan <- sapply(genes_pan, function(x)
                range(chrY_subseg$dna_segs[[2]]
                      [chrY_subseg$dna_segs[[2]]$gene == x,])
                 )
annot_pan <- annotation(x1 = x_pan[1,], x2 = x_pan[2,],
                        text = dimnames(x_pan)[[2]])
```

The segments can be directly plotted, passing the annotations as a list.

```{r ex5_plot, fig.width=7, fig.height=2}
main <- "Comparison of two subsegments in H. sapiens and P. troglodytes"
plot_gene_map(chrY_subseg$dna_segs, chrY_subseg$comparison,
              annotations = list(annot_homo, annot_pan),
              dna_seg_scale = TRUE,
              main = main,
              scale = FALSE)
```

## Example 6: Combining several **genoPlotR** figures and annotating the figure

In this example, some of the previous plots are combined in a single, 
multi-panel figure. Using the tools present in the `grid` package, the
plot is annotated further. More information is available in the documentation
of the `grid` package.

The example here uses **R** objects built in the previous examples, which 
should thus be run first.

First, a `viewport` that will contain all plots is pushed in a new page.
The plot is divided in three rows, of relative heights 1, 1.3 and 0.8.

```{r ex6_vpext_fake, eval=FALSE}
pushViewport(viewport(
  layout = grid.layout(3, 1,
                     heights = unit(c(1,1.3,0.8), rep("null", 3))),
  name = "overall_vp"))
```

The three panels A to C, containing the result of the above examples 4 to 6 
are pushed into viewports, calling `upViewport()` after each plot to 
come back to the main viewport. After the last plot, `upViewport(0)`
is called to go up 2 viewports, coming back to the root viewport. Each time,
the argument `plot_new = FALSE` is used, to avoid plotting in a new 
page every single panel.

```{r ex6_plot_fake, eval=FALSE}
## Panel A
pushViewport(viewport(layout.pos.row = 1, name = "panelA"))
plot_gene_map(dna_segs = bbone$dna_segs,
              comparisons = bbone$comparisons,
              dna_seg_scale = c(FALSE, FALSE, FALSE, TRUE),
              scale = FALSE,
              main = "A", main_pos = "left",
              plot_new = FALSE)
upViewport()
## Panel B
pushViewport(viewport(layout.pos.row = 2, name = "panelB"))
plot_gene_map(barto$dna_segs, barto$comparisons, 
              annotations = annots,
              tree = tree_barto,
              xlims = xlims,
              limit_to_longest_dna_seg = FALSE,
              scale = FALSE,
              dna_seg_scale = TRUE,
              main = "B", main_pos = "left",
              annotation_height = 0.6, annotation_cex = 0.5, 
              plot_new = FALSE)
upViewport()
## Panel C
pushViewport(viewport(layout.pos.row = 3, name = "panelC"))
plot_gene_map(chrY_subseg$dna_segs, chrY_subseg$comparison,
              annotations = list(annot_homo, annot_pan),
              dna_seg_scale = TRUE, scale = FALSE,
              main = "C", main_pos = "left",
              plot_new = FALSE)
upViewport(0)
```

```{r ex6_plot, echo=FALSE, fig.width=7, fig.height=8}
pushViewport(viewport(
  layout = grid.layout(3, 1,
                     heights = unit(c(1,1.3,0.8), rep("null", 3))),
  name = "overall_vp"))

## Panel A
pushViewport(viewport(layout.pos.row = 1, name = "panelA"))
plot_gene_map(dna_segs = bbone$dna_segs,
              comparisons = bbone$comparisons,
              dna_seg_scale = c(FALSE, FALSE, FALSE, TRUE),
              scale = FALSE,
              main = "A", main_pos = "left",
              plot_new = FALSE)
upViewport()
## Panel B
pushViewport(viewport(layout.pos.row = 2, name = "panelB"))
plot_gene_map(barto$dna_segs, barto$comparisons, 
              annotations = annots,
              tree = tree_barto,
              xlims = xlims,
              limit_to_longest_dna_seg = FALSE,
              scale = FALSE,
              dna_seg_scale = TRUE,
              main = "B", main_pos = "left",
              annotation_height = 0.6, annotation_cex = 0.5, 
              plot_new = FALSE)
upViewport()
## Panel C
pushViewport(viewport(layout.pos.row = 3, name = "panelC"))
plot_gene_map(chrY_subseg$dna_segs, chrY_subseg$comparison,
              annotations = list(annot_homo, annot_pan),
              dna_seg_scale = TRUE, scale = FALSE,
              main = "C", main_pos = "left",
              plot_new = FALSE)
upViewport(0)
```

The functions `current.vpTree()` and `grid.ls()` can be used to see the current 
viewport structure, and to see all object names.

```{r ex6_currentvp_fake, eval=FALSE}
grid_list <- grid.ls(grob = TRUE, viewports = TRUE, print = FALSE)
str(grid_list)
current.vpTree()
```

```{r ex6_currentvp, echo=FALSE, fig.show='hide'}
pushViewport(viewport(
  layout = grid.layout(3, 1,
                     heights = unit(c(1,1.3,0.8), rep("null", 3))),
  name = "overall_vp"))

## Panel A
pushViewport(viewport(layout.pos.row = 1, name = "panelA"))
plot_gene_map(dna_segs = bbone$dna_segs,
              comparisons = bbone$comparisons,
              dna_seg_scale = c(FALSE, FALSE, FALSE, TRUE),
              scale = FALSE,
              main = "A", main_pos = "left",
              plot_new = FALSE)
upViewport()
## Panel B
pushViewport(viewport(layout.pos.row = 2, name = "panelB"))
plot_gene_map(barto$dna_segs, barto$comparisons, 
              annotations = annots,
              tree = tree_barto,
              xlims = xlims,
              limit_to_longest_dna_seg = FALSE,
              scale = FALSE,
              dna_seg_scale = TRUE,
              main = "B", main_pos = "left",
              annotation_height = 0.6, annotation_cex = 0.5, 
              plot_new = FALSE)
upViewport()
## Panel C
pushViewport(viewport(layout.pos.row = 3, name = "panelC"))
plot_gene_map(chrY_subseg$dna_segs, chrY_subseg$comparison,
              annotations = list(annot_homo, annot_pan),
              dna_seg_scale = TRUE, scale = FALSE,
              main = "C", main_pos = "left",
              plot_new = FALSE)
upViewport(0)

grid_list <- grid.ls(grob = TRUE, viewports = TRUE, print = FALSE)
str(grid_list)
current.vpTree()
```

Despite the complexity of the viewport structure, it is possible to identify
the previously defined panel A, B and C. Using the `downViewport()` 
function, it will be possible to find and modify elements of the plots. 

First, the labels of panel A will be modified to have the names in italics
and to replace the underscores by a dot and a space. The second label is 
also removed. Note the use of `grid.edit()` and `grid.remove()`.

Second, a red rectangle will be placed around one subsegment of
panel B.

The plot is shown before and after these 2 edits.

```{r ex6_mod_panela_fake, eval=FALSE}
downViewport("panelA")
for (i in 1:length(names)) {
  new_label <- sub("_", ". ", names[[i]])
  grid.edit(paste("label", i, sep = "."),
            label = new_label, 
            gp = gpar(fontface = "italic"),
            redraw = FALSE)
}
grid.remove("label.2")
upViewport(0)
```

```{r ex6_mod_panelb_fake, eval=FALSE}
downViewport("panelB")
downViewport("dna_seg.3.2")
grid.rect(height = unit(2.2, "npc"), gp = gpar(col = "red", lwd = 2, fill = 0))
upViewport(0)
```

```{r ex6_mod_panela, echo=FALSE, fig.width=7, fig.height=8}
pushViewport(viewport(
  layout = grid.layout(3, 1,
                     heights = unit(c(1,1.3,0.8), rep("null", 3))),
  name = "overall_vp"))

## Panel A
pushViewport(viewport(layout.pos.row = 1, name = "panelA"))
plot_gene_map(dna_segs = bbone$dna_segs,
              comparisons = bbone$comparisons,
              dna_seg_scale = c(FALSE, FALSE, FALSE, TRUE),
              scale = FALSE,
              main = "A", main_pos = "left",
              plot_new = FALSE)
upViewport()
## Panel B
pushViewport(viewport(layout.pos.row = 2, name = "panelB"))
plot_gene_map(barto$dna_segs, barto$comparisons, 
              annotations = annots,
              tree = tree_barto,
              xlims = xlims,
              limit_to_longest_dna_seg = FALSE,
              scale = FALSE,
              dna_seg_scale = TRUE,
              main = "B", main_pos = "left",
              annotation_height = 0.6, annotation_cex = 0.5, 
              plot_new = FALSE)
upViewport()
## Panel C
pushViewport(viewport(layout.pos.row = 3, name = "panelC"))
plot_gene_map(chrY_subseg$dna_segs, chrY_subseg$comparison,
              annotations = list(annot_homo, annot_pan),
              dna_seg_scale = TRUE, scale = FALSE,
              main = "C", main_pos = "left",
              plot_new = FALSE)
upViewport(0)


downViewport("panelA")
for (i in 1:length(names)) {
  new_label <- sub("_", ". ", names[[i]])
  grid.edit(paste("label", i, sep = "."),
            label = new_label, 
            gp = gpar(fontface = "italic"),
            redraw = FALSE)
}
grid.remove("label.2")
upViewport(0)


downViewport("panelB")
downViewport("dna_seg.3.2")
grid.rect(height = unit(2.2, "npc"), gp = gpar(col = "red", lwd = 2, fill = 0))
upViewport(0)

```

## Example 7: Using user-defined gene_type and seg_plot

In this example, the use of personalized `gene_types` and of 
`seg_plots` to show data next to DNA segments is demonstrated. The 
same data as in example 4 is used.

A graphical function is defined, to represent triangles instead of arrows 
for genes. The triangles are oriented in function of their strand and the 
transparency is set differently according to a length cut-off, passed with 
the function. As required, the function has only two arguments, a `gene` 
which is a row of a `dna_seg`, and the dots (`...`). Any extra argument 
passed to `plot_gene_map()` will be passed in turn to the 
`gene_type`-defining function.

```{r ex7_starGrob, results='hide'}
starGrob <- function(gene, ...) {
  ## Coordinates for the star
  x <- sin(((0:5) / 2.5) * pi) * (gene$end - gene$start) / 2 + (gene$end + gene$start) / 2
  y <- cos(((0:5) / 2.5) * pi) * gene$strand * 0.5 + 0.5
  idx <- c(1, 3, 5, 2, 4, 1)
  ## Attribute line_col only if present in the gene
  line_col <- if (!is.null(gene$line_col)) gene$line_col else gene$col
  ## Having a conditional transparency, depending on a length cut-off
  ## passed via dots
  length_cutoff <- list(...)$length_cutoff
  if (!is.null(length_cutoff)) {
    alpha <- if ((gene$end-gene$start) < length_cutoff) 0.3 else 0.8
  } else {
    alpha <- 1
  }
  
  ## Grobs
  g <- polygonGrob(x[idx], y[idx],
                   gp = gpar(fill = gene$col, col = line_col,
                             lty = gene$lty, lwd = gene$lwd, alpha = alpha),
                   default.units = "native")
  t <- textGrob(label = "***",
                x = (gene$end+gene$start) / 2,
                y = 0.5,
                default.units = "native")
  gList(g, t)
}
```

To make use of that, the `gene_type` column of the `dna_segs` must be replaced. 
To make use of the supplementary argument `line_col`, a column is added to the 
`dna_seg` object. Finally, plot the result.

```{r ex7_replace_geneType, fig.width=7, fig.height=6}
barto$dna_segs[[2]]$gene_type <- "starGrob"
barto$dna_segs[[4]]$gene_type <- "starGrob"
line_col <- rep(1:20, (nrow(barto$dna_segs[[3]]) %% 20) + 1)
barto$dna_segs[[2]]$line_col <- line_col[1:nrow(barto$dna_segs[[2]])]
plot_gene_map(barto$dna_segs, barto$comparisons,
              tree = tree_barto,
              annotations = annots,
              xlims = xlims, 
              dna_seg_scale = TRUE,
              length_cutoff = 600,
              annotation_height = 0.3)
```

Now, let's plot some random data along with the DNA segment. A standard 
`grid` function returning a grob is used, but it is possible to use 
personalized functions, as long as they return `grob` or `gList`
objects. For example, the function `starGrob()` defined in the previous 
exercise could be modified to be used here.

The `seg_plot` objects are first defined, with the following structure:

```{r ex7_definesegplots}
seg_plots <- lapply(barto$dna_segs, function(ds) {
  x <- seq(1, range(ds)[2], by = 1000)
  y <- jitter(seq(100, 300, length = length(x)), amount = 50)
  seg_plot(func = linesGrob, args = list(x = x, y = y,
                                         gp = gpar(col = grey(0.3), lty = 2)))
})
str(seg_plots[[1]])
```

Each `seg_plot` element has a `func` argument, giving the function 
to use to plot the data, here the `grid` function `linesGrob()`. The
arguments that are passed to this function are contained in the `args` 
argument, a list. Notice the use of `gp = gpar(...)`. 

The result can then be plotted, and modulated using the arguments 
`seg_plot_height`, `seg_plot_height_unit`, `seg_plot_yaxis`,
and `seg_plot_yaxis_cex`.

```{r ex7_plotwithsegplots, fig.width=7, fig.height=6}
plot_gene_map(barto$dna_segs, barto$comparisons, tree = tree_barto,
              annotations = annots,
              xlims = xlims,
              seg_plots = seg_plots,
              seg_plot_height = 0.5,
              seg_plot_height_unit = "null",
              seg_plot_yaxis = 2,
              seg_plot_yaxis_cex = 0.7)
```
